---
title: "Identifying Beginning of WR Motion"
author: "Nolan Birkeland"
date: "`r Sys.Date()`"
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading Packages and Data

```{r}
suppressMessages(library(tidyverse))
data_dir <- "../data"

suppressMessages({
  plays <- read_csv(file.path(data_dir, "plays.csv"))
  players <- read_csv(file.path(data_dir, "players.csv"))
  player_plays <- read_csv(file.path(data_dir, "player_play.csv"))
  tracking <- read_csv(file.path(data_dir, "tracking_week_1.csv"))
})
```

## Cleaning

```{r}
# Only tracking WRs
receivers <- players %>% filter(position == 'WR')

# Only week 1 games
gameIds <- unique(tracking$gameId)

# Only plays with WRs
motion_plays <- player_plays %>%
  filter(motionSinceLineset == TRUE & gameId %in% gameIds & nflId %in% receivers$nflId)

head(motion_plays)
```

Going forward, we will only use the first play as an example.

```{r}
gid <- motion_plays$gameId[1]
pid <- motion_plays$playId[1]
receiver <- motion_plays$nflId[1]

play <- tracking %>% filter(gameId == gid & playId == pid & nflId == receiver)
```

## Basic Plotting

```{r}
snap <- play %>% filter(frameType == "SNAP") %>%
  pull(frameId) %>%
  first()

play %>%
  ggplot(aes(x = frameId, y = s)) +
  geom_line() +
  geom_line(aes(x = frameId, y = a), color = "blue") + 
  geom_vline(xintercept = snap, color = "red")
```

We can clearly see a large spike in both speed and acceleration (s and a) at around frame 60. This is likely the beginning of the WR motion. Additionally, we can see from both the speed and acceleration curve that the WR in this example, Stefon Diggs, comes to a stop before the snap.

## Identifying Beginning of Motion

To identify the beginning of motion, we will use a spline curve to smooth the speed curve. 

```{r}
library(splines)

spline_fit <- smooth.spline(play$frameId, play$s)
plot(play$frameId, play$s, xlab = "Frame", ylab = "Speed")
lines(spline_fit, col = "blue", lwd = 2)
```

The curve fits the path perfectly. Now, we can take the first derivative (acceleration), and the second (jerk).

```{r}
speed <- predict(spline_fit, x = play$frameId)$y
accel <- predict(spline_fit, x = play$frameId, deriv = 1)$y
jerk <- predict(spline_fit, x = play$frameId, deriv = 2)$y
```

Using the assumption that the beginning of motion is the point where the WR is accelerating most rapidly before the snap, we can find the frame where jerk is maximized.

```{r}
# Remove frames at/after snap
jerk <- jerk[1:snap-1]
max_jerk <- which.max(jerk)

print(max_jerk)
```

This indicates that at frame 57, Stephon Diggs begins his motion.

```{r}
play %>%
  ggplot(aes(x = frameId, y = s)) +
  geom_line() +
  geom_line(aes(x = frameId, y = a), color = "blue") + 
  geom_vline(xintercept = snap, color = "red") +
  geom_vline(xintercept = max_jerk, color = "green")
```

We can see that this line just about perfectly aligns with the start of motion.

## Another Example

We will demonstrate this approach with a play featuring motion from Gabe Davis.

```{r}
gid <- motion_plays$gameId[2]
pid <- motion_plays$playId[2]
receiver <- motion_plays$nflId[2]

play <- tracking %>% filter(gameId == gid & playId == pid & nflId == receiver)
```

### Plotting

```{r}
snap <- play %>% filter(frameType == "SNAP") %>%
  pull(frameId) %>%
  first()

play %>%
  ggplot(aes(x = frameId, y = s)) +
  geom_line() +
  geom_line(aes(x = frameId, y = a), color = "blue") + 
  geom_vline(xintercept = snap, color = "red")
```


### Spline Curve

```{r}
spline_fit <- smooth.spline(play$frameId, play$s)
plot(play$frameId, play$s, xlab = "Frame", ylab = "Speed")
lines(spline_fit, col = "blue", lwd = 2)

speed <- predict(spline_fit, x = play$frameId)$y
accel <- predict(spline_fit, x = play$frameId, deriv = 1)$y
jerk <- predict(spline_fit, x = play$frameId, deriv = 2)$y

# Remove frames at/after snap
jerk <- jerk[1:snap-1]
max_jerk <- which.max(jerk)

print(max_jerk)
```

### Results

```{r}
play %>%
  ggplot(aes(x = frameId, y = s)) +
  geom_line() +
  geom_line(aes(x = frameId, y = a), color = "blue") + 
  geom_vline(xintercept = snap, color = "red") +
  geom_vline(xintercept = max_jerk, color = "green")
```

